---
title: "Wallenius' noncentral multivariate hypergeometrical distribution"
output: html_document
---

## Biased sampling

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

Today I needed to better understand what the resulting distribution is, when sampling without replacement using the R `sample` function, where each element exists exactly once and has a different weight of being drawn. Let $M$ be the total number of elements (i.e. colors when using urn terminology) and $N$ the number of elements sampled from $M$, and $P$ a vector of weights. Repeatedly sampling $N=2$ out of $M=4$ with $P=[4,3,2,1]$ gives the mean occurence of each element in the sample.

```{r}
N <- 2
M <- 4
P <- M:1
reps <- 1e4
v <- replicate(reps, sample(1:M, size=N, replace=FALSE, prob=P))
tbl <- table(unlist(v))
tbl / reps
```

The distribution that models this setting is [Wallenius' noncentral multivariate hypergeometrical distribution (WNMHD)](https://en.wikipedia.org/wiki/Wallenius%27_noncentral_hypergeometric_distribution). Conceptually, it is the same as the multivariate hypergeometrical distribution except for that the elements additionally have weights. The probability for a certain result can be calculated using the corresponding density function from the package `BiasedUrn`. For example, drawing the elements ${1,3}$ for the same parameters as above yields the probability

```{r}
library(BiasedUrn)
dMWNCHypergeo(x = c(1,0,1,0), m = rep(1, M), n=N, odds=P)
```

Repeating this process for every possible outcome for a given $M$ and $N$, we can calculate the distribution of means for the WNMHD.

```{r}
# approach one: calculate all combinations
l <- replicate(M, 0:1, simplify = F)  # all combinations 
d <- t(as.matrix(expand.grid(l)))     # generate all combinations
d <- d[ , colSums(d) == N]            # columns with N elements drawn
p <- dMWNCHypergeo(x = d, m = rep(1, M), n=N, odds=P)  # get density
mu <- as.vector(d %*% cbind(p))       # calculating mean
mu 
```

Using this brute force approach quickly runs into trouble when $M$ gets large. The calculation of all possible combinations will quickly fail due to memory limits. A better way to calculate the means is to use function `meanMWNCHypergeo`. Settings the `precision` argument to a small value will yield an exact solution.

```{r}
mu2 <- meanMWNCHypergeo(m = rep(1, M), n=N, odds = P, precision = .0)
mu2
```

Additionally, the function also allows to estimate the means with reduced precision for high values of $M$ and $N$. By this one can evade the trouble caused by the brute force approach above.

```{r}
M <- 20
N <- 10
P <- M:1
mu2 <- meanMWNCHypergeo(m = rep(1, M), n=N, odds = P, precision = .1)
mu2
```



## Two stage sampling strategy

Using this knowledge, we can calculate the resulting distribution for a sampling strategy applied in my research. In my setting, $N$ is not fixed but is also a random variable used as a hyper-parameter in the second sampling stage using WNMHD as described above. As all draws are indendent, the expected means of the compound distribution can simply be calculated as a weighted sum of the estimated WNMHD means for every possible realization of $N$. Let the distribution for the random variable $N$ be  

```{r}
P1 <- c("1" = .25, "2"=.5, "3"=.25) 
P1
```

Calculating the weighted sum of estimated WNMHD means for each N gives.

```{r}
# stage 2 sampling
M <- 4
P <- M:1
MU <- mapply(meanMWNCHypergeo, n=1:3, 
            MoreArgs = list(m = rep(1, M), odds = P, precision = 0))
MU <- t(MU)
P1 %*% MU
```

To see if the results are correct let's try it again using a simulation.

```{r}
# stage 1 sampling
reps <- 1e5
N1 <- sample(1:3, reps, replace=T, prob=P1)

v <- lapply(N1, function(x)  {
  sample(1:M, size=x, replace=FALSE, prob=P)
})
tbl <- table(unlist(v))
tbl / reps
```

The results are almost identical. As the WNMHD means are estimated and thus only approximately correct, we cannot expect identity of the results even for many replications. 
